---
title: "More EDA"
output: html_notebook
---

You should have already run the "WS_week2_EDA.Rmd" worksheet and become familiar with working with an RStudio RMarkdown document.

Here we're going to go a little further and explore a new dataset, the [KDD99](http://kdd.ics.uci.edu/databases/kddcup99/kddcup99.html) dataset. Read about the competition [task specification](http://kdd.ics.uci.edu/databases/kddcup99/task.html).

Go get the [10%](http://kdd.ics.uci.edu/databases/kddcup99/kddcup.data_10_percent.gz) and [column names](http://kdd.ics.uci.edu/databases/kddcup99/kddcup.names) data from there. They are read in as follows:

```{r}
kddata<-read.csv("../data/kddcup.data_10_percent.gz")
kddnames=read.table("../data/kddcup.names",sep=":",skip=1,as.is=T)
colnames(kddata)=c(kddnames[,1],"normal")
```


Lets take a look:
```{r}
head(kddata)
```
It is marked up as red team/blue team. Everything is numeric except the "normal" column which contains the truth labels about the process that generated the data.
```{r}
summary(kddata)
```

Q1. Make a barplot of the "normal" column, using an appropriate scaling. Consider using "par(las=2)". Use the inline help (?par) to see how this is done and what it does. Consider the scaling of the "y" axis.

```{r}
par(las=2)
barplot((sort(table(kddata[,"normal"]))),log="y")
```

Now we'll examine the labels separately. This is a way to make a list for each class:
```{r}
labs=unique(as.character(kddata[,"normal"]))
names(labs)=labs
kddlist=lapply(labs,function(x){
  kddata[kddata[,"normal"]==x,1:41]
})
```

Q2. What does the function colMeans do? Use it and sapply to make a matrix of the mean of each *numeric* variable, within each group, and plot the result as a *heatmap*. 

What are the 7 most important features, in terms of mean differences? Don't standardize rows or columns, but consider *scaling* the whole matrix for visualisation.

List them in a variable called "mycols" for future use.
```{r}
kddmean=t(sapply(kddlist,function(x)colMeans(x[,c(1,5:41)])))
library("gplots")
heatmap.2(log(kddmean+1),margins =c(9,15),trace="none")
mycols=c("dst_bytes","src_bytes","duration","dst_host_svd_count","dst_host_count","srv_count","count")
```
Q2b. Now standardize the features and repeat the *heatmap*.
Do the important features change? What does this mean about the solvability of the problem?
```{r}
kddfreq=apply(kddmean,2,function(x)x/(sum(x)+1))
heatmap.2(kddfreq[,!is.nan(kddfreq[1,])],margins =c(9,15),trace="none")
```

Q3. Make a *table* of the interaction between the class label and the *categorical* variables. 

Describe some key features that you see in the data: are any of the classes *informative*? In what sense are they informative?
```{r}
mycategorical=colnames(kddata)[2:4]
classlist=lapply(mycategorical,function(mycat){
  table(kddata[,c(mycat,"normal")])
})
for(i in 1:3) heatmap.2(log(classlist[[i]]+1),margins =c(9,15),trace="none",main=mycategorical[i])
```
Q4. Learn about the *variability* in the labels with the function "sd", given the mean. 

*Represent* that in an appropriate format. Again, what interpretations can you make about the data?
```{r}
kddsd=t(sapply(kddlist,function(x){
  apply(x[,c(1,5:41)],2,sd)
}))
heatmap.2(log(kddsd/(kddmean+0.01)+1),margins =c(9,15),trace="none")
```
Q5. The task specifically notes that the dataset contains attacks that are not listed here. 

Comment on what the above Exploratory Data analysis might mean for the hopes of detecting different properties of attack, and how you might go about making a model that will perform well out-of-sample, *comparing* "normal" to other classes of "attack".


